---
title: "Constructing drug episodes for pancreatic (gemcitabine)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing drug episodes for pancreatic (gemcitabine)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(episodes)
library(dplyr)
library(lubridate)
library(purrr)
```

## Sample Size Breakdown

```{r echo = FALSE}
# Drug episodes for lung
drug_episodes <- prep_episode_data(tumour = "pancreatic", treatment = "gemcitabine", drug_separator = ",",
                              overlap_threshold = 1)

state_transitions <- construct_state_episodes(drug_episodes)

drug_transitions <- state_transitions$drug_transitions
```

## Descriptive statistics of episodes stratified by line and state

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(lubridate)


# Step 1: Prepare data
summary_data <- drug_transitions %>%
  arrange(patientid, linestartdate) %>%
  group_by(patientid) %>%
  mutate(
    state_clean = state,
    next_state = lead(state_clean),
    died = any(state_clean == "Death"),
    is_last = row_number() == n(),
    
    # Safe death_censor assignment
    death_censor = if_else(!is.na(next_state) & next_state == "Death", 1, 0),
    
    # Admin censor if last row and not died and not going to next line
    admin_censor = if_else(
      is_last & !died & state_clean != "On_Treatment_Next_Line", 
      1, 0
    ),
    
    duration_days = as.numeric(lineenddate - linestartdate)
  ) %>%
  ungroup() %>%
  filter(!(state_clean %in% c("Death", "On_Treatment_Next_Line")))  
 

# Step 2: Updated summary function
summarize_group <- function(df) {
  tibble(
    N = length(unique(df$patientid)),
    Mean = round(mean(df$duration_days, na.rm = TRUE), 1),
    Median = round(median(df$duration_days, na.rm = TRUE), 1),
    Q1 = round(quantile(df$duration_days, 0.25, na.rm = TRUE), 1),
    Q3 = round(quantile(df$duration_days, 0.75, na.rm = TRUE), 1),
    `Death Censoring` = paste0(sum(df$death_censor), " (", round(100 * mean(df$death_censor), 1), "%)"),
    `Admin Censoring` = paste0(sum(df$admin_censor), " (", round(100 * mean(df$admin_censor), 1), "%)")
  )
}

summary_by_line <- summary_data %>%
  group_by(chemo_line, state) %>%
  group_split() %>%
  purrr::map_df(~ summarize_group(.x) %>%
                  mutate(Group = paste0("Line ", unique(.x$chemo_line), " - ", unique(.x$state))),
                .id = NULL)

# Step 3b: Overall summary by state (not stratified by line)
summary_overall <- summary_data %>%
  group_by(state) %>%
  group_split() %>%
  purrr::map_df(~ summarize_group(.x) %>%
                  mutate(Group = paste0("Overall - ", unique(.x$state))),
                .id = NULL)

# Step 3c: Combine both
summary_table <- bind_rows(summary_by_line, summary_overall) %>%
  select(Group, everything())


knitr::kable(summary_table, caption = "Summary of Episode Duration by Line and State (Excl. Death State)")
```

# Cut-points 

## Cumulative incidence functions of transitions out of each health state

## Exit from on-treatment

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(tidycmprsk)
library(ggplot2)
library(survival)
library(ggsurvfit)
library(cmprsk)

# Step 1: Prepare dataset to define transitions
transitions <- drug_transitions %>%
  arrange(patientid, linestartdate) %>%
  group_by(patientid) %>%
  mutate(
    from_state = state,
    from_date = linestartdate,
    to_state = lead(state),
    to_date = lead(linestartdate),
    time_to_transition = as.numeric(to_date - from_date),
    status = ifelse(!is.na(to_state), 1, 0)
  ) %>%
  ungroup() %>%
  filter(!is.na(to_state))  # only rows with an actual transition

# Step 2: Restrict to specific origin states (as requested)
valid_from_states <- c("On_Treatment_Target_Line", "off_treatment", "progression")
transitions <- transitions %>%
  filter(from_state %in% valid_from_states)

# Step 3: Clean up data and handle edge cases
transitions <- transitions %>%
  filter(
    time_to_transition > 0,  # Remove negative or zero transition times
    !is.na(time_to_transition)  # Remove missing transition times
  ) %>%
  mutate(
    # Ensure to_state is a factor with proper levels
    to_state = factor(to_state),
    from_state = factor(from_state, levels = valid_from_states)
  )

# Check if we have data after filtering
if(nrow(transitions) == 0) {
  stop("No valid transitions found after filtering")
}

# Step 4: Create survival object and compute CIF
# Note: For competing risks, event_code should start from 1 (not 0)
transitions <- transitions %>%
  mutate(
    event_code = as.numeric(from_state)  # This will be 1, 2, 3, etc.
  )

cif_tidy <- transitions %>%
  tidycmprsk::cuminc(Surv(time_to_transition, from_state) ~ to_state, data = .)



# Step 1: Filter only from On_Treatment_Target_Line
transitions_filtered <- transitions %>%
  filter(from_state == "On_Treatment_Target_Line") %>%
  mutate(
    event_code = dplyr::case_when(
      to_state == "off_treatment" ~ 1,
      to_state == "progression" ~ 2,
      to_state == "Death" ~ 3,
      TRUE ~ 0  # Censored
    )
  )

# Step 2: Run competing risks model
cif <- cmprsk::cuminc(
  ftime = transitions_filtered$time_to_transition,
  fstatus = transitions_filtered$event_code
)

# Step 3: Plot it
plot(cif, col = c("tomato", "steelblue", "darkgreen"), lwd = 3)
legend("topright", legend = c("Off-Treatment", "Progression", "Death"), 
       col = c("tomato", "steelblue", "darkgreen"), lwd = 4)



```

## Exit from progression

```{r echo = FALSE, message = FALSE, warning = FALSE}
ggcuminc(cif_tidy, outcome = "progression", size = 1.5) +
  scale_color_manual(values = c("steelblue", "darkgreen", "tomato", "purple")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  ) +
  guides(color = guide_legend(override.aes = list(linewidth = 2)))
```

## Exit from Off-treatment

```{r echo = FALSE, message = FALSE, warning = FALSE}
ggcuminc(cif_tidy, outcome = "off_treatment", size = 1.5) +
  scale_color_manual(values = c("steelblue", "darkgreen", "tomato", "purple")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  ) +
  guides(color = guide_legend(override.aes = list(linewidth = 2)))
```

## Restricted mean survival time survival: comparisons across lines

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(survRM2)

os_data <- drug_transitions %>%
  arrange(linestartdate) %>%
  group_by(patientid) %>%
  mutate(event = if_else(row_number() == n() & state == "Death", 1L, 0L)) %>%
  summarise(linestartdate = min(linestartdate), lineenddate = max(lineenddate), event = max(event), chemo_line = first(chemo_line)) %>%
  mutate(time = as.numeric(lineenddate - linestartdate))

# Prep the data
rmst_data <- os_data %>%
  mutate(
    time = as.numeric(lineenddate - linestartdate),
    status = event
  ) %>%
  filter(!is.na(time), time >= 0)

tau_80 <- quantile(rmst_data$time[rmst_data$status == 1], 0.8, na.rm = TRUE)


unique_lines <- unique(rmst_data$chemo_line)
comparisons <- list()

for(i in 1:(length(unique_lines)-1)) {
  for(j in (i+1):length(unique_lines)) {
    line_i <- unique_lines[i]
    line_j <- unique_lines[j]
    
    comparison_data <- rmst_data %>%
      filter(chemo_line %in% c(line_i, line_j)) %>%
      mutate(arm = if_else(chemo_line == line_i, 0, 1))
    
    if(nrow(comparison_data) > 0) {
      result <- rmst2(
        time = comparison_data$time,
        status = comparison_data$status,
        arm = comparison_data$arm,
        tau = tau_80
      )
      
      comparisons[[paste(line_i, "vs", line_j)]] <- list(
        comparison = paste(line_i, "vs", line_j),
        rmst_diff = result$unadjusted.result[1,1],
        ci_lower = result$unadjusted.result[1,2],
        ci_upper = result$unadjusted.result[1,3],
        p_value = result$unadjusted.result[1,4],
        rmst_line1 = result$RMST.arm0$rmst[1],
        rmst_line2 = result$RMST.arm1$rmst[1]
      )
    }
  }
}

# Convert to a nice table
comparison_table <- map_dfr(comparisons, ~ tibble(
  Comparison = .x$comparison,
  RMST_Difference = round(.x$rmst_diff, 2),
  CI_Lower = round(.x$ci_lower, 2),
  CI_Upper = round(.x$ci_upper, 2),
  P_Value = round(.x$p_value, 4),
  RMST_Group1 = round(.x$rmst_line1, 2),
  RMST_Group2 = round(.x$rmst_line2, 2)
))


knitr::kable(
  comparison_table,
  caption = "Restricted Mean Survival Time Comparisons Between Lines of Therapy",
  col.names = c(
    "Comparison",
    "RMST Difference (days)",
    "95% CI Lower",
    "95% CI Upper",
    "P-Value",
    "RMST Group 1",
    "RMST Group 2"
  ),
  align = "lrrrrrr",
  format = "markdown"  
)

```

## KM curve: time to death (overall survival)

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(survminer)

km_fit_overall <- survfit(Surv(time, event) ~ 1, data = os_data)

ggsurvplot(
  km_fit_overall,
  conf.int = TRUE,
  risk.table = TRUE,
  title = "Overall Survival Curve",
  xlab = "Days Since Line Start",
  ylab = "Survival Probability",
  surv.median.line = "hv",
  ggtheme = theme_minimal()
)
```

## KM curve: time to death (stratified by line)

```{r echo = FALSE, message = FALSE, warning = FALSE}
km_fit_strat <- survfit(Surv(time, event) ~ factor(chemo_line), data = os_data)

ggsurvplot(
  km_fit_strat,
  conf.int = TRUE,
  risk.table = TRUE,
  title = "Survival Stratified by Line of Therapy",
  xlab = "Days Since Line Start",
  ylab = "Survival Probability",
  surv.median.line = "hv",
  ggtheme = theme_minimal(),
  legend.title = "Chemo Line",
  legend.labs = paste("Line", sort(unique(os_data$chemo_line)))
)

```

