---
title: "Constructing drug episodes and deriving Weibull transition estimates: Assumptions and Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing drug episodes and deriving Weibull transition estimates: Assumptions and Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(episodes)
library(dplyr)
library(lubridate)
```


## Walkthrough of package functions and assumptions

## Prepping the episode data

The first function in the package workflow is `prep_episode_data`. This is used to clean up the tumour-specific Flatiron drug episodes datasets. This dataset finally outputs the cleaned and merged drug episodes for only 1st, 2nd and 3rd line for the tumour-specific patients who have received the treatment of interest. 

The function can take in a regular so for example, for lung we match to either cisplatin or carboplatin (as they are often interchanged in the protocols). `exact = TRUE` can be specified to match exactly that treatment without any combination of other drugs. For lung, we specify `exact = FALSE` as cisplatin and carobplatin is rarely given alone. However, for all ohter tumours for which we have produced drug episodes we match exactly to the treatment name. 


```{r}
drug_episodes <- prep_episode_data(tumour = "nsclc", treatment = "cisplatin|carboplatin", drug_separator = ",",
                              overlap_threshold = 1, exact = FALSE)

head(drug_episodes)
```

## Overlapping lines 

The `prep_episode_data` function wraps a function to deal with overlapping lines where the `linestartdate` and `lineenddate` of the previous and next line overlapped. An episode was flagged as overlapping if its start date preceded the end date of the prior episode.  

Contiguous overlapping lines were then grouped using a cumulative indexing strategy. Specifically, a new group identifier was initiated each time a non-overlapping line was encountered, while overlapping lines shared the same identifier. Within each group, start and end dates were recalculated to span the earliest and latest dates, respectively, and all unique drug names were combined into a single standardized line label. Duplicate or redundant drug entries were cleaned using regular expressions to ensure consistency.Finally, the collapsed episodes were renumbered sequentially for each patient to reflect the revised line structure.

```{r}
# Sample overlapping input
example_data <- tibble::tibble(
  patientid = c(1, 1, 1),
  linenumber = c(1, 2, 3),
  linename = c("cisplatin", "cisplatin", "gemcitabine"),
  linestartdate = lubridate::ymd(c("2020-01-01", "2020-01-15", "2020-02-01")),
  lineenddate = lubridate::ymd(c("2020-01-30", "2020-02-15", "2020-03-01"))
)

example_data

# Collapse overlapping lines
collapsed_data <- collapse_overlapping_lines(example_data)

print(collapsed_data)
```

## Clean up single doses defined as lines 

In the Flatiron dataset, we found several lines where the `linestartdate` was equal to the `lineenddate`. These are likely single doses which were not administered for a period of time, and thus these should not be categorized as a line of therapy. The `single_dose_clean` function aims to clean or collapse such single-day episodes by either: 1. Merging them into temporally adjacent lines with overlapping drug names, or 2. Dropping them if they appear spurious and disconnected. The `drug_separator` parameter can be used to specify how a combination drug is written in the dataset e.g. if the name is `cisplatin, gemcitabine`, the parameter is set to `,`.

First, all treatment episodes are chronologically ordered and flagged as single-day lines if their `linestartdate` and `lineenddate` are equal. These lines are then evaluated in the context of their immediate neighbors (preceding and following lines) within the same patient.

A merging criterion is applied whereby a single-day line is considered for consolidation if it shares one drug components with a temporally adjacent line. This overlap can be adjusted by the `overlap_threshold` parameter (default is 1). Overlap is assessed at the level of individual drug names, allowing for flexibility in combination regimens. If this overlap criterion is met, the single-day line is merged into the adjacent episode, with the resulting line defined by the earliest start date, the latest end date, and a deduplicated, alphabetically ordered list of drug components.

Single-day lines that fail to meet the merging criterion—i.e., those that are both isolated and non-overlapping with adjacent lines—are flagged for removal, under the assumption that they may reflect data entry inconsistencies or non-line of therapeutic events.


```{r}
# Sample data
example_data <- tibble::tibble(
  patientid = c(1, 1, 1),
  linenumber = c(1, 2, 3),
  linename = c("cisplatin", "cisplatin", "gemcitabine"),
  linestartdate = ymd(c("2022-01-01", "2022-01-15", "2022-01-15")),
  lineenddate   = ymd(c("2022-01-10", "2022-01-15", "2022-01-15"))  # line 3 is single-day
)

example_data

# Clean single-day lines
cleaned_data <- single_dose_clean(example_data)

print(cleaned_data)
```

## Time period restriction 

We found that in the Flatiron dataset, very few patients started treatment before 2010, with the significant bulk of the treatment start date mass centered after 2010. Given the advancements and differences in treatments, as well as the issue of these patients biasing survival times by having an unusually long tail, we filtered the dataset for each tumour to patients only after 2010. 

## Cutoffs for time to death, time to progression etc. 

In terms of cutoffs for what would be considered a valid time to death from on-treatment, time to progression from on-treatment, time to next line from on-treatment and time to off-treatment,  we decided to use the 75 \% quantile as a cutoff. This ensures that the cutoff is tumour-specific. This cutoff point was supported by other statistical measures, namely a log-rank test which often showed a separation in groups around the Q3 point, a Classification and Regression Tree (CART) whose first split point also was consistent with the Q3 and a Guassian Mixture clustering model, which often clustered observations beyond Q3 together. 

## Progression from on-treatment

In the Flatiron dataset, we have several progression dates available for a single patient depending on when they received their scans. For a valid on-treatment to progression state transition for a particular treatment, we only consider progression if it occurs atleast 7 days after the start of line-treatment and atleast within 2-weeks of the line end date. As we only model transitions from the target line, any transition to a new line is also considered progression. 

## Death Date granularity

In the Flatiron dataset, we only have the month and year of death for the patient and not the exact date. Thus, to infer an exact date, we take the maximum date of the date of last followup, the progression date, the date of last clinical note, and the date of treatment end depending on what information is available. If all these dates are not available, the date is inffered to be 15th, for approximately the middle of the month.

## Constructing the drug episodes

The `construct_state_episodes` function can be used to construct the drug episodes into a form which can be fed into the Weibull models to derive the sojourn survival time. It takes in the cleaned output from `prep_episodes_data`.

```{r}
state_episodes <- construct_state_episodes(drug_episodes)

drug_transitions <- state_episodes$drug_transitions

lines <- state_episodes$lines
```

## Deriving the Weibull shape and scale parameters

To derive the shape and scale parameters, we use a Weibull model with the accelerated failure time parameterization from the `flexsurv` package. We derive shape and scale parameters for exit from on-treatment, exit from off-treatment and exit from progression. To use the `state_exit_weibull_estimates` function, we pass in the `drug_episodes` and `lines` datasets, both of which are outputs from the `construct_state_episodes` function. 

In all three regressions, we adjust for the number of prior lines (unless all the patients have no prior lines). For the exit from progression, we also have a binary indicator variable (`LoT_progression` with teo levels `LoT` and `clinical`) to control for the fact that progression was derived for some patients clinically through the progression date and inferred for others through a change in line. This is to control for the artifact that patients that were proxied to be progressed by a change in line of therapy will likely stay in the progression state for a longer period of time.

```{r}
params_results <- state_exit_weibull_estimates(drug_transitions, lines)

params_results
```

